---
title: "Class 12 Pt. 2"
author: "Kyle Canturia (A17502778)"
format: pdf
---

# Section 1: Identify genetic variants of interest

Downloaded data from Ensembl as a csv file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

GBR dataset

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

# Section 4: Population Scale Analysis

```{r}
pop <- read.table("rs8067378_ENSG00000172057.6.txt")
head(pop)
#reads dataset and assigns it to a variable
```

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
table(pop$geno)
##only looks at the column "geno" and returns count of how many of each genotype is present
```
Theres 108 A|A, 233 A|G, and 121 G|G.
```{r}
library(dplyr) 
  pop %>%
  group_by(geno) %>%
  summarize(geno_median = median(exp))
##uses dplyr tools to first group the samples by genotype, then summarizes the median of each genotype
```

The median expression levels for A|A, A|G, and G|G are 31.24, 25.06, and 20.07 respectively. 

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
ggplot(pop) + 
  aes(geno, exp) +
  geom_boxplot()
##makes boxplot with geno on the x axis and exp on the y
```

A|A has the highest expression value while G|G has the lowest, and A|G is in between. From the plot, it's evident that the SNP is associated with differing expressions of ORMDL3.





